# Nonsense or no senses?

## Introduction

In linguistic terms, clouds may provide us with different types of information, both at syntagmatic and paradigmatic level. At the syntagmatic level, they may illustrate cases of collocation, colligation, semantic preference or even tendencies towards the open-choice principle. The paradigmatic level, on the other hand, codes the relationship between the clusters and dictionary senses, from heterogeneous clusters to those that represent prototypical contexts of a sense.

With a naive understanding of the correlation between context and meaning, we would mostly expect, from the paradigmatic perspective, clusters that equal senses: each cluster would cover all the occurrences of a dictionary sense and only the occurrences of that sense. However, even if we relax the requirements, expecting clusters with mostly occurrences of one sense and covering most of the instances that are not discarded as noise, this does not happen often. Instead, even homogeneous clusters only group typical contexts within a sense, which, at the syntagmatic level, tend to correspond to collocations. In any case, the full picture is much more complex, and we may obtain much richer information than just lexical collocations representing typical contexts within a sense.

In this chapter, we will look into the types of syntagmatic and paradigmatic information that the clouds offer. Section \@ref(infotypes) starts with an overview of the different levels in each dimension and mentions a few examples of their interaction in a contingency table. We then elaborate with more detailed examples of each in situation in sections \@ref(collocation) through \@ref(openchoice), and round up with an overall summary in Section \@ref(summary).

## Types of information {#infotypes}

The linguistic information obtainable from the clusters can be understood from the syntagmatic perspective as co-occurrence patterns of different kinds, and from the paradigmatic perspective in relation to dictionary senses. Both dimensions interlace, resulting in a number of specific phenomena that we may encounter. The relationship is summarized in Table \@ref(tab:colsem); the syntagmatic or collocational dimension is represented by the columns and discussed in Section \@ref(collocationally), and the paradigmatic or semantic dimension is represented by the rows and discussed in Section \@ref(semantically).

```{r, colsem, anchor = 'Table'}
read_tsv(
  # here::here("assets", "collocation-prototypicality-contingency_latex.txt"),
  here::here("assets", "collocation-prototypicality-contingency_html.txt"),
  col_types = cols()
) %>% 
  knitr::kable(booktabs = T, escape = F,
               caption = "Contingency table between the collocational and semantic perspectives, with a few examples.",
               linesep = "\\addlinespace") %>% 
  kable_styling(full_width = T) %>% 
  column_spec(1, width = "6em", bold = T) %>% 
  column_spec(2:5, width = "8.5em")
```

### Collocational perspective {#collocationally}

In order to interpret the different levels of information that a syntagmatic or collocational perspective may offer us, we can make use of some theoretical concepts from the foundations of Corpus Linguistics. @sinclair_1998 [124-125] established a framework for collocational analysis that included, next to the node, i.e. our targets, four key components The obligatory component, semantic prosody, will not be taken into consideration here. The other three will help us make sense of the observed output of the clouds. These are: collocation, colligation and semantic preference.

In their simplest form, collocations are defined as the co-occurrence of two words within a certain span [@firth_1957a 13; @sinclair_1991 170; @sinclair_1998 15; @stubbs_2009 124]. They might be further refined by the statistical significance of their co-occurrence frequency or by their strength of attraction (e.g. \textsc{pmi}; see @mcenery.hardie_2012 [122-133] for a discussion). Even though a collocational relationship is asymmetric, that is, the co-occurrence with a more frequent word B may be more important for the less frequent word A than for B, the measures used to described it are most often symmetrical.
<!-- NOTE maybe add Gries' paper? -->
In any case, their relevance to the interpretation of clouds takes a different form, definitely asymmetric. Considering models built around a target term or node, frequent, distinct context words are bound to make the tokens that co-occur with them similar to each other and different from the rest: they will generate clusters. In terms of clouds, these are found as clusters defined by one context word or a group of co-occurring context words with a high F-score towards the cluster. In other words, most of the tokens in a cluster co-occur with the same word(s), and most of the tokens co-occurring with these words are in that cluster.
<!-- These context words tend to have a high PMI with the target. -->
In some cases, two frequent context words co-occur, and clusters may be defined by whether the tokens co-occur with both of them or only one.

In contrast to collocation, understood as a relationship between words (and, traditionally, as a relationship between word forms), colligation is defined as a relationship between grammatical categories or between a word and grammatical categories [@firth_1957a 14; @sinclair_1998 15; @stubbs_2009 124]. In order to capture proper colligations, we would need models in which parts of speech or maybe dependency patterns are used as features, which is not the case in these studies. However, by rejecting a strict separation between syntax and lexis,
<!-- TODO find references for this? -->
we can make a grammatically-oriented intepretation of collocations with function words, such as frequent prepositions, the *om te* combination, or the passive auxiliary. Given this caveat, we will talk about lexically instantiated colligations when we encounter clusters dominated by items that indicate a specific grammatical function.

Semantic preference is defined as the relationship between a word and semantically similar words [@sinclair_1998 16; @stubbs_2009 125; @mcenery.hardie_2012 138-140]. Within traditional collocational studies, this implies grouping collocates, that is, already frequently co-occurring items, based on semantic similarity. Its identification requires more interpretation on the part of the researcher, but token-level models provide a more objective grounding in terms of type-level distances, as well as the possibility to detect semantic patterns across infrequent co-occurrences. For our purposes, semantic preference is translated as clusters that are not dominated by a single collocate or group of co-occurring collocates, but are instead defined by a group of infrequent context words with similar type-level vectors and for which we can give a semantic interpretation. (Cases of similar context words without a semantic interpretation are quite rare, and normally involve pronouns or adverbs.)

The three notions described above assume identifiable patterns: occurrences that repeat themselves often enough, and are different enough from other contexts, that their instantiations form a cluster. Going back to @sinclair_1991's founding notions, we are assuming the domination of the idiom principle:

> ...a language user has available to him or her a large number of semi-preconstructed phrases that constitute single choices, even though they might appear to be analysable into segments. [@sinclair_1991 110]

The opposite situation would be given by the open-choice principle:

> At each point where a unit is completed (a word or a phrase or a clause), a large range of choice opens up and the only restraint is grammaticalness. [@sinclair_1991 109]

Both notions, if understood as poles in the continuum of collocational behaviour, can help us interpret the variety of shapes that we encounter within and across lemmas. Lemmas that tend to find identifiable clusters, with strong collocations, lexically instantiated colligations or sets with semantic preference, can be said to tend towards the idiom principle. In contrast, lemmas that tend to exhibit large proportions of noise tokens, and small, diffuse clusters (Cirrus clouds, mostly), can be said to tend towards the open-choice principle. With this reasoning, next to the three categories described above, we include the tendency towards open-choice as a fourth category, meant to include the clouds that do not conform to either of the idiom-principle formats.

### Semantic perspective {#semantically}

In terms of the relationship between the HDBSCAN clusters and the manually annotated dictionary senses, we can initially distinguished between heterogeneous clusters, i.e. where the cluster does not exhibit a clear preference for one sense, and homogeneous clusters. Secondly, the homogeneous clusters may cover all the (clustered) tokens of a given sense, or only part of it, i.e. a (proto)typical context of the sense. Additionally, said (proto)typical context may highlight a certain aspect or dimension of the meaning of the target, different from that highlighted by a different context.

As a result, the semantic dimension covers four different types of situations. The first one, i.e. heterogeneous clusters or clusters with multiple senses, would normally be interpreted as bad modelling, if we consider the senses a gold standard and the target of our models. It is also the most frequent interpretation of the open-choice clouds. Nonetheless, they can also occur in other kinds of clouds, and as such illustrate the mismatch between contextual and semantic structure: clear contextual patterns do not imply dictionary senses. The second type of situation, i.e. clusters that match senses, is the normal expectation from a naive point of view that truly equates patterns to dictionary cases. It is quite rare and often indicative of fixed expressions or very particular meanings.

Rather than full senses, contextual patterns tend to represent typical contexts of a sense. 
The notion of prototypicality in Cognitive Semantics is related to the perspective that categories need not be discrete and uniform, but instead may exhibit fuzzy boundaries and/or degrees of membership. This at the extensional level; it also works at the intensional level, but we will not get into that right now. If we define a sense based on a set of overlapping features, not all features need to co-occur for an item to be assigned the sense, but the more features are present, the more central/salient the item is; the better example it is. [This sucks, I have to rephrase.]
As they appear in the clouds, a sense may exhibit one typical context that is much more frequent and clear that the rest, or multiple typical contexts with various frequencies. Unfortunately, neither t-SNE nor HDBSCAN provide a reliable mapping between quantitative properties and relative centrality *of* the clusters. In contrast, we can identify central cases within an HDBSCAN cluster based on their membership probability, which is a normalized *eps* value per cluster. Items with a higher membership probability lie in a denser area of an HDBSCAN cluster, and therefore have more items similar to it than the items in sparser areas. They do not necessarily occur in the euclidean centre in the t-SNE plot, but might form one or more dense cores closer towards an edge instead. In addition, we can distinguish between rather uniform clusters, in which all members have a similar weight, from more diverse clusters with dense cores and sparse peripheries.

Extensional prototypicality works at all levels. We could identify (proto)typical instances/contexts of a lemma, of a sense, or of a dimension of a sense. For the sake of illustration, let's take frequency as a proxy for saliency or prototypicality, and look at one medoid of *heffen*. The five most frequent context words (the only ones co-occurring with more than 10 tokens) are, in descending order: *nemen*, *journalist*, *toerist*, *meter* and *worden*. From this point of view, *nemen*, which represents a rather fixed expression *een horde nemen* 'to take over an obstacle', is the single most typical context of the lemma *heffen*. It is not the most typical sense, however, because both *journalist* and *toerist* represent the 'horde' sense, and their combined frequencies are slightly larger than that of *nemen*. Next, within a sense, both *nemen* and *meter* represent the typical contexts of their respective senses, i.e. metaphorical and literal 'hurdle', whereas *journalist* and *toerist* both instantiate the same sense 'horde'. Neither of them can be said to be semantically central to the 'horde' meaning, given their rather similar frequencies. They are just two typical contexts for the same sense. But in this case we can also go deeper, by distinguishing different aspects of the 'horde' sense. The rest of the context words co-occurring with the *journalist* and *toerist* tokens point towards different dimensions, so that actually *journalist* is the typical context for the 'intimidating, dangerous or violent mass of people' dimension, while *toerist* is the typical context for the 'massive flow of people' dimension. Intensionally, each of these dimensions have a predominant weight in their respective clusters.

### Interaction between dimensions

As we can see in Table \@ref(tab:colsem), the interaction between the four levels of each dimension result in a 4x4 table with all but one cell filled with at least one example. Naturally, not all the combinations are equally frequent or interesting: the most salient one is certainly the collocation that indicates the prototypical context of a sense. But this does not mean that the rest of the phenomena should be ignored: we can still find interesting and useful information with other shapes of clouds, other contextual patterns, other semantic structure.

In the following sections, we will look in detail at examples of each attested combination. Each section will focus on one level of the collocational dimension, and will be further subdivided by the levels of the semantic dimension.
The examples will be illustrated with plots in which the colours represent HDBSCAN clusters, the shapes indicate manually annotated dictionary senses, and the transparency, the *eps* value from HDBSCAN. The senses are not specified in the legends, while the clusters will be named with the context word that represents the most and its F-score. Concretely, for each cluster we calculate the precision and recall of the context words that co-occur with its tokens, i.e. the proportion of tokens in the cluster that co-occur with the word, as well as the proportion of tokens co-occurring with the word that are included in the cluster. From these values, an F-score is calculated: the higher it is, the better a cue the context word is for the cluster. The highest value, 1, indicates that all the tokens co-occurring with that word, and only the tokens co-occurring with that word, are within that cluster. The context word with the highest F-score, and that F-score, are used to name the clusters in these plots.

## Collocation {#collocation}

The first level of the collocational or syntagmatic dimension is that of the collocation: clusters dominated by one context word or a group of co-occurring context words. They are most likely to be found as Cumulus clouds, but also as Stratocumulus clouds or, vary rarely, Cirrus clouds.

### Heterogeneous clouds

Albeit infrequently, collocations might transcend senses, that is, they might be frequent and even distinctive of a lemma without showing a preference for a specific sense.
The most clear example is *heilzaam*, whose clusters are typically dominated by one context word that is not indicative of any one sense, as shown in Figure \@ref(fig:heilzaam). The most typical collocates are *werking* and *effect* (adding, in some models, the less frequent *invloed*), which tend to occur in attributive constructions (*de heilzame werking van X* or *X heeft een heilzame werking*, which in fact form distinct clusters in another model). Among the rest of the tokens, the most relevant collocate is the copula *zijn*, joined by *werken*, which indicate a predicative construction, over the most frequent attributive one, and therefore belong to the category of lexically instantiated colligations.

(ref:heilzaam) Cloud of *heilzaam*.

```{r, heilzaam, fig.cap = '(ref:heilzaam)'}
plotWithCws('heilzaam', 3)
```

## One sense

In a few cases we can see clusters characterized by one context word (or a couple of always co-occurring context words) that perfectly match a sense (or at least its clustered tokens). These are normally fixed expressions, at least to a degree.

The best example is shown in Figure \@ref(fig:schaal), where a cluster characterized by *Richter* and one characterized by *gewicht* represent the 'range' sense of *schaal* and its idiomatic sense (the dish of a metaphorical scale) respectively.

It is, however, a bit nuanced. First, there are a few (3 or 4) tokens of the 'range' sense that are not covered by *Richter* and indeed have a very different pattern: instead of *de schaal van X*, they occur as *een schaal van A tot B*. These are the orange dots at the top of the plot. In addition, *Richter* tends to co-occur with *de* and *van*, but neither of those context words are as distinctive of the sense, so when (in other models) the cluster includes tokens co-occurring with *van* but not with *Richter*, it becomes less homogeneous.

Second, *gewicht* typically co-occurs with *leggen*, but can also co-occur with *werpen*. This cluster, also in other models, tends to work around *gewicht*, normally excluding tokens that only co-occur with *leggen*, which do not belong to the same idiom any more.

(ref:schaal) Cloud of *schaal*.

```{r, schaal, fig.cap = '(ref:schaal)'}
plotWithCws('schaal', 1)
```

### (Proto)typical context

The most frequent phenomenon --- especially among Cumulus clouds --- is a cluster dominated by one context word or group of co-occurring context words that represents a (proto)typical context of a sense. It may be *the* prototypical context, if the rest of the the sense is discarded as noise or spread around less clear clusters, but we might also find multiple clusters representing different typical contexts of the same sense. Neither t-SNE nor HDBSCAN can tell whether one of these contexts is more central than the other, although a rough approximation can be achieved by recalculating the membership probability scores for groups of clusters.

The most clear example is *heffen*, whose typical objects are also characteristic of its two main senses (see Figure \@ref(fig:heffen)). On the one hand, the 'to levy' sense occurs mostly with *belasting*, *tol*, and *accijns*; their frequencies seem to be large enough to form three distinct clusters, but they are linked in the following level of the HDBSCAN hierarchy. On the other hand, the 'to lift' sense occurs with *glas*, where the final expression takes a metonymical meaning 'to toast', and with the body-parts *hand*, *arm* and *vinger*, in which they might take other metonymical meanings. The latter does not really belong to the this "collocation" category but to "semantic preference".

In terms of frequency, the *belasting* cluster --- which includes similar tokens that do not co-occur with *belasting* --- is the most prominent one. It can also be said to dominate the 'to levy' sense, although, based on the *eps* values provided by HDBSCAN, the tokens co-occurring with *accijns*, followed by those co-occurring with *tol*, are more central to the combined 'to levy' cluster than those co-occurring with *belasting* in this model.

(ref:heffen) Cloud of *heffen*.

```{r, heffen, fig.cap = '(ref:heffen)'}
plotWithCws('heffen', 1)
```

It must be noted that these clusters may not be just characterized by one single context word, but by multiple co-occurring context words, and the clusters may split to distinguish contexts that have all or some of them. A clear example is *hachelijk* (Figure \@ref(fig:hachelijk)), where both senses are characterized by prototypical contexts: *onderneming*, *zaak* and *avontuur* for the 'potentially dangerous' sense, *moment*, *situatie*, and *positie* for the 'critical' sense. These six frequent context words are paradigmatic alternatives of each other, all taking the slot of the modified noun. However, unlike its very similar (at type-level) alternative *situatie*, *positie* may also co-occur with *bevrijden* (and *uit*) and, additionally, with *brandweer*, typically in Belgian contexts. The frequency of these co-occurrences in the sample, next to the type-level dissimilarity between these three lexical items, splits the co-occurrences with *positie* in three clusters, based on these combinations.

It would be hard to argue about the relative centrality of the three *positie* clusters. In a way, they result from the combination of three features, and each cluster exhibits a different degree of membership based on how many of these overlapping features it co-occurs with. At the same time, they have a distinctive regional distribution. Based on this data, we might said that a prototypical context of *hachelijke posities* 'dangerous/critical positions' in Flanders is a situation in which firefighters free someone/something from them, while this core is not present, or at least not nearly as relevant, in the Netherlands. We might also say that the same situation is not typical of *hachelijke situaties* 'dangerous/critical situations', and this therefore presents a (local) distributional difference between two types that otherwise, at corpus level, are very similar.

(ref:hachelijk) Cloud of *hachelijk*.

```{r, hachelijk, fig.cap = '(ref:hachelijk)'}
plotWithCws('hachelijk', 1)
```

### Profiling

Clusters dominated by a context word may not only represent a (proto)typical context within a sense, but also one that highlights a different dimension of such sense than other clusters. This is not extremely frequent and requires an extra layer of interpretation, but it is quite straightforward to look for.

<!-- For example, *tol*, *accijns* and *belasting* apply to different contexts, in the sense that they are literally applied to different circumstances: roads in the case of *tol*, and (typically) fuel in the case of *accijns*. However, from the point of view of the target *heffen*, this doesn't really imply a different meaning. -->

One example is given by the 'substance' meaning of *stof*, represented as circles in Figure \@ref(fig:stof).
Within this sense, we tend to find clusters dominated by *gevaarlijk*, *schadelijk* (which also attracts *kankerwekkend*) and *giftig* (which often attracts *chemisch*, and sometimes is separated from it). These dominant context words are not very different from each other at type-level, and indeed, their clusters belong to the same branch in the HDBSCAN hierarchy.

However, we do find slightly different themes among the less frequent context words in each of those clusters, which correspond to different situations or dimensions: the environment and composition of other substances for *schadelijk* and *kankerwekkend* (*uitstoot*, *lucht*, *stank*, *bevatten*); and the context of drugs or as something that is liberated for *giftig* and *chemisch* (*vormen*, *vrijkomen*, *drugsbebruik*).
This effect of the less frequent context words is one of the consequences of "noisier" models: at some levels of analysis, one word (*gevaarlijk*, *schadelijk*...) might be enough to disambiguate the target, but this extra information enriches our understanding of how the words are actually used. It is also contextualized information: not just about how *stof* is used and what it means, but how it is used (and what it means) when in combination with certain frequent collocates.

(ref:stof) Cloud of *stof*.

```{r, stof, fig.cap = '(ref:stof)'}
plotWithCws('stof', 3)
```

## Lexically instantiated colligation

Some defined by the presence of a word that indicates a grammatical preference or tendency (valency?), such as passive constructions.

### Heterogeneous

The senses of *herstructureren* emerge from a combination of specialization (whether it's specifically applied to companies) and argument structure (intransitive or transitive constructions). The most frequent lexical collocates, *bedrijf* and *grondig*, rarely co-occur with each other or with *worden*, which indicates passive construction and therefore a transitive form. The clusters in various models tend to be built around these three context words, grouping most of the transitive uses, either company-specific or not, in the *worden* clusters, such as the blue one in Figure \@ref(fig:herstructureren).

(ref:herstructureren) Cloud of *herstructureren*.

```{r, herstructureren, fig.cap = '(ref:herstructureren)'}
plotWithCws('herstructureren', 6)
```

### One sense

While a rare thing, we might be able to pinpoint a specific sense based on a collocation that points to a grammatical pattern. One clear case is the reflexive sense of *herhalen*, characterized by its co-occurrence with *zich* (Figure \@ref(fig:herhalen)), which is not captured by `FOC-POS:lex` models. Even when it is, it might further be split by the co-occurrence with *geschiedenis* in about half the attestations.

We expected this result in other lemmas with purely reflexive senses as well, but it is not easy to achieve. For example, *herinneren* is often split in three major clusters with a small one for *eraan* that is sometimes absorbed by a larger cloud. The large ones are characterized by *aan*, *ik* and *me*, and *zich*, respectively. Both the first and third person are frequent enough within the reflexive sense to give rise to their own clusters. In the case of *diskwalificeren*, instead, the reflexive sense is typically (but not always) absorbed within the transitive sense that matches it semantically, i.e. the non sports-related sense.

(ref:herhalen) Cloud of *herhalen*.

```{r, herhalen, fig.cap = '(ref:herhalen)'}
plotWithCws('herhalen', 7)
```

### (Proto)typical context

The most clear example of clusters indicating typical contexts of a sense and defined by grammatical collocates is given by *diskwalificeren* and the tendency of the sports-specific sense to co-occur with *worden*. This is shown by the yellow cluster in Figure \@ref(fig:diskwalificeren).

(ref:diskwalificeren) Cloud of *diskwalificeren*.

```{r, diskwalificeren, fig.cap = '(ref:diskwalificeren)'}
plotWithCws('diskwalificeren', 4)
```

### Profiling

The 'horde' sense of *horde* tends to co-occur with *tourist* and *journalist* in this corpus. The two collocates are quite similar to each other, but the rest of the context words in their clusters point towards a different dimension of the 'horde' sense: hordes of journalists/photographers will surround and follow celebrities (*omringen*, *opwachten*, *achtervolgen*), while hordes of tourists will instead flood and move around in the city (*toestromen*, *stad*). This is strengthened by the co-occurrence of the preposition *door* and the verb *worden* with *journalist* --- to the point that the corresponding cluster is better represented by *door* than by *journalist* in Figure \@ref(fig:horde) ---, while *toerist* prefers *naar*.

(ref:horde) Cloud of *horde*.

```{r, horde, fig.cap = '(ref:horde)'}
plotWithCws('horde', 4)
```

## Semantic preference

Clusters defined by multiple infrequent similar context words.

### Heterogeneous

Just like we can have clusters dominated by one context word that is not characteristic of one sense, we can have clusters dominated by multiple similar context words that are not characteristic of any sense. This is the case of names of colours and clothes, which (somehow) includes *haar* 'hair'. As a result, *grijs* tokens pointing to the general grey things and to, specifically, grey/white hairs, are clustered together (the large light blue cluster in Figure \@ref(fig:grijs)). Note that, visually, the two senses occupy opposite halves of this cluster.

A similar group of context words is responsible for joining the 'fabric' and 'lit. dust' senses of *stof*, even across homonyms.

(ref:grijs) Cloud of *grijs*.

```{r, grijs, fig.cap = '(ref:grijs)'}
plotWithCws('grijs', 4)
```

Another case is the set of juridical terms in *herroepen*: the most frequent of them, *uitspraak*, is ambiguous: when it means 'statement', *herroepen* means 'to retract', whereas its 'verdict' sense invokes the 'to annul' sense of *herroepen*. Both situations occur in the sample, but the most frequent one is the 'statement' sense, that is, not the one of the juridical field. In some models, the *uitspraak* tokens form their own cluster, next to another cluster with context words such as *rechtbank*, *vonnis*, etc., in which case the former is heterogeneous and the latter, homogeneous. In other models (see Figure \@ref(fig:herroepen)), both groups are clustered together, leading to a heterogeneous cluster brought together by semantic preference, although again, visually they are distinguishable. At the same time, *verklaring* and *bekentenis* also could be considered part of the same semantic field, in broad terms, but they point to a different aspect, i.e. a different frame within the same field of legal action, and, correspondingly, their type-level vectors are different and they tend to represent distinct, homogeneous clusters.

(ref:herroepen) Cloud of *herroepen*.

```{r, herroepen, fig.cap = '(ref:herroepen)'}
plotWithCws('herroepen', 6)
```

### One sense

A few senses can be completely clustered by groups of similar context words, as long as we don't mind too much the portion of the sense that was excluded as noise. One of these cases is the same group of *schaal* tokens represented by *Richter*, which, in models that exclude *Richter*, can be grouped by *kracht*, *aardbeving* and related context words. Another example is found in *haken*, where the 'make someone trip' sense is characterized by football-related terms (*strafschop(sgebied)*, *penalty*, *scheidsrechter*...), and the very infrequent 'crochet' sense, by *breien*, *naaien*, *hobby* and similar. They are represented in the dark blue and light blue clusters in Figure \@ref(fig:haken) respectively.

(ref:haken) Cloud of *haken*.

```{r, haken, fig.cap = '(ref:haken)'}
plotWithCws('haken', 1)
```

### (Proto)typical context

There are several examples of clusters defined by semantically similar infrequent context words, representing typical contexts of a sense. In Figure \@ref(fig:grijs), for example, the dark blue cluster is represented by cars, mostly thanks to *Mercedes* and *Opel*, but also other brands. In the case of some lemmas, particularly *dof*, some models split the tokens co-occurring with different related context words in multiple clusters (*klinken*, *knal*, *klap*, *dreun*) while others group them together in one large cluster defined by a semantic preference indicative of a sense (in this case, 'dull (sound)').

A typical semantic group is the culinary one: found with *schaal* 'dish' and with *heet* (the red cluster in Figure \@ref(fig:heet)). In the case of *heet*, almost all the tokens co-occurring in this cluster refer to literally hot foods and drinks, although the full expression might be idiomatic, i.e. *de soep niet zo heet eten*, and only a few of them belong to the much less frequent sense 'spicy'. In other models, either the *soep + eten* tokens or the *water* tokens might form separate clusters. In addition, *aardappel* is very similar to the other context words in this group at the type-level, but it still tends to form its own cluster, thanks to both its frequency and the distinctiveness of its larger cotext.

(ref:heet) Cloud of *heet*.

```{r, heet, fig.cap = '(ref:heet)'}
plotWithCws('heet', 3) +
  scale_shape_manual(values = c(16, 17, 18, 0, 2, 4, 8), guide = "none")
```

### Profiling

Among the senses of *geldig*, the specific one related to regulated validity is much more frequent than the general one. In addition, models tend to distinguish two dimensions of this regulated validity, either as clusters characterized by semantic preference, or simply as regions in the t-SNE plot. One of the groups represents contexts in which someone has to present (*voorleggen*, *bezitten*...) some form of identification (*rijbewijs*, *paspoort*, etc.), while the other represents other kinds of documents (*ticket*, *abonnement*) in which the duration of its validity is more salient (*tot*, *maand*, numbers...). While it is not clear from their most representative context words, these groups are illustrated as the green and yellow clusters, respectively, in Figure \@ref(fig:geldig).


(ref:geldig) Cloud of *geldig*.

```{r, geldig, fig.cap = '(ref:geldig)'}
plotWithCws('geldig', 1)
```

## Towards open-choice {#openchoice}

Most Cirrus clouds, as well as the Cumulonimbus, are not really grouped by one dominant context word or by a set of similar words. It is not clear what brings the tokens together. Naturally, this mostly leads to heterogeneous clouds, but occasionally we might find more homogeneous ones among them. The principle behind this category is not so much that the clusters are bad, but that the lemmas seem to follow more of an open-choice principle: a wider variety of options, so that the most powerful collocates barely group a couple of tokens. By definition they aren't patterns, so they cannot represent prototypical contexts of a sense, but some of these clusters do happen to be quite homogeneous.

### Heterogeneous

The most common situation in clusters that are not explained by a dominant context word or semantic preference, especially when they are Cumulonimbus, is that they are semantically heterogeneous. One such example is the Cumulonimbus (the large cloud) of *helpen* in Figure \@ref(fig:helpen), formed in opposition to the smaller Stratocumulus, which is also an heterogeneous cloud formed by semantic preference for kinship terms.

(ref:helpen) Cloud of *helpen*.

```{r, helpen, fig.cap = '(ref:helpen)'}
plotWithCws('helpen', 1)
```

### One sense

Cumulonimbus clouds are a sort of bag with 'all the rest' after some powerful context words have defined what it takes to make a cluster. Typically, they are heterogeneous. However, in the case of *huldigen*, the 'to hold (an opinion)' sense has strong patterns and the 'to pay homage' sense covers the rest of the tokens. In Figure \@ref(fig:huldigen) we can see, next to the small Cumulus and Stratocumulus characterized by *principe*, *standpunt* and *opvatting* (typical of the 'to hold (an opinion)' sense), a large Cumulonimbus. It has a dense core (hail) made of tokens that only co-occur with *worden* and a larger, more diffuse area characterized by different combinations of *worden*, sports-related terms (*kampioen*, *winnaar*, *sportsraad*) as well as town administration terms (*stadsbestuur*, *gemeentebestuur*...). These do not form one semantic field, but instead come from different areas that co-occur in the context of 'to pay homage': ceremonies organized by sports and city organizations in public places, in honour of successful athletes.

(ref:huldigen) Cloud of *huldigen*.

```{r, huldigen, fig.cap = '(ref:huldigen)'}
plotWithCws('huldigen', 6)
```

### Part of a sense

In models that tend to have diffuse clusters, we might have a cluster that holds itself together for unclear reasons and yet represents a part of a sense fairly well. This is certainly not a frequent co-occurrence, but it might indicate that there is a sense-related organization in space, that is however not strong enough to stand out as a defined cluster.

In our data, we find this in the "best" medoid of *hoop*, where the 'heap' tokens, united by *een*, form a rather diffuse island, slightly separated from the 'hope' tokens (Figure \@ref(fig:hoop)). In spite of this visually clear separation made by t-SNE, HDBSCAN only manages to capture patches in the form of weak clusters. One such cluster captures all the 'heap' tokens that are clustered, but a number of them are lost as noise. Because of this, *een* has a perfect recall for this cluster but a very low precision; meanwhile, the rest of the context words involved form no pattern at all.

(ref:hoop) Cloud of *hoop*.

```{r, hoop, fig.cap = '(ref:hoop)'}
plotWithCws('hoop', 8)
```

## Summary